---
title: "Results"
output: html_notebook
author: Guy Lomax
date: 2024-01-12
editor_options: 
  chunk_output_type: console
---

This notebook compares the results of the RPI methodology with those of RUE,
RESTREND and LGS.

```{r setup}

library(tidyverse)
library(terra)
library(sf)
library(here)
library(tmap)

```


```{r load}

# Country polygons
ke_tz <- st_read(here("data", "raw", "vector", "kenya_tanzania.geojson"))

# Potential and actual GPP values from RPI model
potential_gpp_all <- rpi_paths %>%
  map(function(path) rast(path) %>% subset("fitted")) %>%
  rast()
names(potential_gpp_all) <- paste0("potential_", 2000:2018)

actual_gpp_all <- rpi_paths %>%
  map(function(path) rast(path) %>% subset("GPP")) %>%
  rast()
names(actual_gpp_all) <- paste0("actual_", 2000:2018)

# RPI layers and trend
rpi_paths <- Sys.glob(here("data", "processed", "raster", "rpi", "rf_rpi_rast_20*.tif"))

rpi_all <- rpi_paths %>%
  map(function(path) rast(path) %>% subset("rpi")) %>%
  rast()
names(rpi_all) <- paste0("rpi_", 2000:2018)

rpi_ts_trend <- rast(here("data", "processed", "raster", "rpi", "rf_rpi_ts_trend.tif"))

# RUE layers and trend
rue_all <- rast(here("data", "processed", "raster", "rue", "rue_all.tif"))
rue_ts_trend <- rast(here("data", "processed", "raster", "rue", "rue_trend.tif"))

# RESTREND residuals and trend
restrend_residuals <- rast(here("data", "processed", "raster", "restrend", "restrend_resids.tif"))[5:23]
restrend <- rast(here("data", "processed", "raster", "restrend", "restrend_results.tif"))

names(restrend_residuals) <- paste0("restrend_resids_", 2000:2018)
names(restrend) <- c("model_yint", "model_slope", "model_rsq", "model_p_value",
                     "resid_slope", "resid_rsq", "resid_p_value")

# Key covariates for comparison/exploration of results
dynamic_covariate_paths <- Sys.glob(here("data", "raw", "raster", "covariateMaps", "dynamicVars*.tif"))
ppt_all <- dynamic_covariate_paths %>%
  map(function(path) rast(path) %>% subset("precipitation")) %>%
  rast()
tMean_all <- dynamic_covariate_paths %>%
  map(function(path) rast(path) %>% subset("tMean")) %>%
  rast()
tree_cover <- rast(here("data", "raw", "raster", "covariateMaps", "staticVars.tif")) %>%
  subset("wriTCFrac") %>%
  crop(ppt_all)
  

```


Now we calculate the R-squared and the weighted mean absolute error per pixel
for each method. How would that work for RUE? It's basically a linear regression
with intercept = 0, so I could just calculate the annual difference from the
expected RUE value each year? 

```{r pixel_wise_error}

# RUE




```



We can compare the performance of our method with those of others on various
criteria. For mean values across space, the key issues are:
- Explaining a greater share of spatial variation with environmental
covariates (R-squared)
- Reduced overall variance?
- Consistency (lack of sharp boundaries)
- Lack of correlation with MAP or other key vars

For trend, we can assess performance through:
- High R-squared for each pixel
- Low mean absolute error (weighted?) for each pixel
- Consistency in Mann-Kendall trend?
- P-values of Theil-Sen slope or permutation trend?
