---
title: "Results"
output: html_notebook
author: Guy Lomax
date: 2024-01-12
editor_options: 
  chunk_output_type: inline
---

This notebook compares the results of the RPI methodology with those of RUE,
RESTREND and LGS.

```{r setup}

library(tidyverse)
library(terra)
library(sf)
library(here)
library(tmap)

```


```{r load}

# Country polygons
ke_tz <- st_read(here("data", "raw", "vector", "kenya_tanzania.geojson"))

# Potential and actual GPP values from RPI model
rpi_paths <- Sys.glob(here("data", "processed", "raster", "rpi", "rf_rpi_20*.tif"))

potential_gpp_all <- rpi_paths %>%
  map(function(path) rast(path) %>% subset("quantile_pred")) %>%
  rast()
names(potential_gpp_all) <- paste0("potential_", 2000:2018)

actual_gpp_all <- rpi_paths %>%
  map(function(path) rast(path) %>% subset("GPP")) %>%
  rast()
names(actual_gpp_all) <- paste0("actual_", 2000:2018)

mean_gpp_all <- rpi_paths %>%
  map(function(path) rast(path) %>% subset("mean_pred")) %>%
  rast()
names(mean_gpp_all) <- paste0("mean_pred_", 2000:2018)

# RPI layers and trend

rpi_all <- rpi_paths %>%
  map(function(path) rast(path) %>% subset("rpi")) %>%
  rast()
names(rpi_all) <- paste0("rpi_", 2000:2018)

rpi_ts_trend <- rast(here("data", "processed", "raster", "rpi", "rf_rpi_ts_trend.tif"))
rpi_kendall <- rast(here("data", "processed", "raster", "rpi", "rf_rpi_kendall_tau_rast.tif"))

# RUE layers and trend
rue_all <- rast(here("data", "processed", "raster", "rue", "rue_all.tif"))
rue_ts_trend <- rast(here("data", "processed", "raster", "rue", "rue_trend.tif"))
rue_kendall <- rast(here("data", "processed", "raster", "rue", "rue_trend_kendall.tif"))

names(rue_all) <- paste0("rue_", 2000:2018)

# RESTREND residuals and trend
restrend_residuals <- rast(here("data", "processed", "raster", "restrend", "restrend_resids.tif"))[[5:23]]
restrend <- rast(here("data", "processed", "raster", "restrend", "restrend_results.tif"))
restrend_kendall <- rast(here("data", "processed", "raster", "restrend", "restrend_kendall.tif"))
restrend_slope <- rast(here("data", "processed", "raster", "restrend", "restrend_resids.tif"))[[1:4]]

names(restrend_residuals) <- paste0("restrend_resids_", 2000:2018)
# names(restrend) <- c("model_yint", "model_slope", "model_rsq", "model_p_value",
#                      "resid_slope", "resid_rsq", "resid_p_value")

# LGS layers
lgs_annual_paths <- Sys.glob(here("data", "processed", "raster", "lgs", "lgs_annual_*.tif"))
lgs_potential_gpp_all <- lgs_annual_paths %>%
  map(function(path) rast(path) %>% subset("potential")) %>%
  rast()
names(lgs_potential_gpp_all) <- paste0("lgs_potential_", 2000:2018)

lgs_mean_gpp_all <- lgs_annual_paths %>%
  map(function(path) rast(path) %>% subset("mean")) %>%
  rast()


# Key covariates for comparison/exploration of results
dynamic_covariate_paths <- Sys.glob(here("data", "raw", "raster", "covariateMaps", "dynamicVars*.tif"))
ppt_all <- dynamic_covariate_paths %>%
  map(function(path) rast(path) %>% subset("precipitation")) %>%
  rast()
tMean_all <- dynamic_covariate_paths %>%
  map(function(path) rast(path) %>% subset("tMean")) %>%
  rast()
tree_cover <- rast(here("data", "raw", "raster", "covariateMaps", "staticVars.tif")) %>%
  subset("wriTCFrac") %>%
  crop(ppt_all)

years <- 2000:2018
names(ppt_all) <- paste0("precipitation_", years)
names(tMean_all) <- paste0("tMean_", years)

```


Now we calculate the R-squared and the weighted mean absolute error per pixel
for each method.

The weighted mean absolute error is the mean absolute difference between actual and
predicted GPP for each year in a pixel. Methods with the lowest pixel-wise MAE
explain the greatest amount of variance as a function of environmental
covariates. Note that the MAE for pixel-wise methods (RUE and RESTREND) are
likely to be lower, since the pixel-wise model only needs to explain variance
over time in each pixel, rather than also explaining spatial variance.

Since the RPI and LGS methods both estimate the 90th percentile of GPP for a
given combination of covariates, we weight the MAE accordingly: positive errors
(actual GPP > potential) are weighted at Q, while negative errors are weighted
at (1-Q). In order to compare equivalently to MAE calculated from RUE and
RESTREND, we must also scale the weighting factors such that the mean weight
of any point is 1. The weighting factors must therefore fulfil the following
conditions:

1. W_1 = (Q / (1 - Q)) * W_2
2. W_1 * (1 - Q) + W_2 * Q = 1

The solutions to this for Q = 0.9 are W_1 = 5 and W_2 = 5 / 9.

```{r pixel_wise_error}

# First, calculate the actual GPP anomaly (difference between annual GPP
# and mean GPP for that pixel)

actual_gpp_anomaly <- actual_gpp_all - mean(actual_gpp_all)

# RUE

rue_mean <- mean(rue_all)

rue_pred <- ppt_all * rue_mean

rue_pred_anomaly <- rue_pred - mean(rue_pred)

gpp_error_rue <- abs(rue_pred_anomaly - actual_gpp_anomaly)

mae_rue <- mean(gpp_error_rue)

# RESTREND

mae_restrend <- mean(abs(restrend_residuals))

# RPI

# gt_weight <- 5
# lte_weight <- 5 / 9

rpi_gpp_anomaly <- mean_gpp_all - mean(mean_gpp_all)

gpp_error_rpi <- rpi_gpp_anomaly - actual_gpp_anomaly

mae_rpi_anom <- mean(abs(gpp_error_rpi))

# abs_weighted_error_rpi <- app(gpp_error_rpi, function(error) {
#   weighted_error <- ifelse(error <= 0,
#                            error * lte_weight,
#                            error * gt_weight)
#   abs_weighted_error <- abs(weighted_error)
# 
#   abs_weighted_error
# })
# 
# mae_rpi <- mean(abs_weighted_error_rpi)

# Best method per pixel by MAE

best_mae <- which.min(c(mae_rue, mae_restrend, mae_rpi_anom))

```


```{r mae_maps}

fill <- tm_shape(ke_tz) +
  tm_fill("grey95")

borders <- tm_shape(ke_tz) +
  tm_borders()

mae_raster_map <- tm_raster(
  col.scale = tm_scale_continuous(
    limits = c(0, 1000),
    values = "OrRd",
    outliers.trunc = c(TRUE, TRUE)
  ),
  col.legend = tm_legend(
    reverse = TRUE,
    title = expression(
      atop("GPP MAE",
           (kg~C~m^-2~y^-1))
    ),
    frame = FALSE
  )
) +
  tm_layout(asp = 3/4,
            frame = FALSE)

rue_mae_map <- fill +
  tm_shape(mae_rue) +
  mae_raster_map +
  borders +
  tm_title("b) RUE")

restrend_mae_map <- fill +
  tm_shape(mae_restrend) +
  mae_raster_map +
  borders +
  tm_title("c) RESTREND")

rpi_mae_map <- fill +
  tm_shape(mae_rpi_anom) +
  mae_raster_map +
  borders +
  tm_title("a) RPI")

best_mae_map <- fill +
  tm_shape(best_mae) +
  tm_raster(
    col.scale = tm_scale_categorical(
      values = "Set1",
      levels = c(1,2,3),
      labels = c("RUE", "RESTREND", "RPI")
    ),
    col.legend = tm_legend(title = "",
                           frame = FALSE)
  ) +
  tm_layout(asp = 3/4, frame = FALSE) +
  tm_title("d) Best method") +
  borders

rue_mae_map
restrend_mae_map
rpi_mae_map

tmap_save(
  rue_mae_map,
  here("results", "figures", "rue_annual_mae_map.png"),
       width = 12, height = 12, units = "cm", dpi = 250
)

tmap_save(
  rpi_mae_map,
  here("results", "figures", "rpi_annual_mae_map.png"),
       width = 12, height = 12, units = "cm", dpi = 250
)

tmap_save(
  restrend_mae_map,
  here("results", "figures", "restrend_annual_mae_map.png"),
       width = 12, height = 12, units = "cm", dpi = 250
)

tmap_save(
  best_mae_map,
  here("results", "figures", "best_mae_map.png"),
       width = 12, height = 12, units = "cm", dpi = 250
)


```

This is an interesting result! It shows that, in seeking to capture both
spatial and temporal variation in GPP, the RPI method does not account for
temporal variations as well as fitting a separate linear regression to each
pixel. This is not hugely surprising, I suppose, since the pixel-wise approach
accounts for local soil, topographic and vegetation factors by definition,
whereas RPI relies on explicitly defining the variables behind spatial variation.

However, this is also what allows RPI to differentiate highly productive from
less productive areas. The local areas of higher MAE in the RPI map could also
correspond to areas with anomalously high or low productivity relative to their
environment.

This all makes me think that MAE is not the appropriate metric, since it
conflates poor model fit with meaningful spatial anomalies in RPI. What I want
is something like R-squared, in which I am trying to determine how much of the
overall *variance* in GPP within a pixel is explained by the model. That way,
if the RPI values are anomalously high, but don't vary very much, we wouldn't
see elevated MAE. We would just be comparing the variance in actual GPP with
variance in predicted GPP.

I'll calculate this next - it might take a bit more work (and would also require
using the mean, rather than quantile).

In addition, I would like to use LGS to generate annual values and compare these,
since this is possible. I would keep the same clusters, though, rather than
allowing them to vary year-on-year. Or should I allow them to vary?


```{r pixel_wise_r_sq}

# Total variance in GPP
actual_gpp_mean <- mean(actual_gpp_all)

actual_gpp_anomaly <- actual_gpp_all - actual_gpp_mean

# Total sum of squares
actual_gpp_tss <- sum(actual_gpp_anomaly ^ 2)

# RUE

rue_rss <- sum(gpp_error_rue ^ 2)

rue_rsq <- 1 - (rue_rss / actual_gpp_tss)

rue_rsq_lt0 <- rue_rsq < 0
rue_rsq_lt0 <- mask(rue_rsq_lt0, rue_rsq_lt0, maskvalues = c(0,NA))

# RESTREND

restrend_rss <- sum(restrend_residuals ^ 2)

restrend_rsq <- 1 - (restrend_rss / actual_gpp_tss)

# RPI

rpi_rss <- sum(gpp_error_rpi ^ 2)

rpi_rsq <- 1 - (rpi_rss / actual_gpp_tss)

rpi_rsq_lt0 <- rpi_rsq < 0
rpi_rsq_lt0 <- mask(rpi_rsq_lt0, rpi_rsq_lt0, maskvalues = c(0, NA))

# Raster of best performing strategy

best_rsq <- which.max(c(rue_rsq, restrend_rsq, rpi_rsq))


```



Visualise R-squared:

```{r pixel_wise_r_sq_viz}

fill <- tm_shape(ke_tz) + 
  tm_fill("grey95")

borders <- tm_shape(ke_tz) +
  tm_borders()

rsq_raster_params <- tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(0, 0.75),
      midpoint = 0,
      values = "PiYG",
      outliers.trunc = c(TRUE, TRUE)
    ),
    col.legend = tm_legend(
      title = expression(R^2),
      reverse = TRUE,
      frame = FALSE
    )
  ) 


rue_rsq_map <- fill +
  tm_shape(rue_rsq) +
  rsq_raster_params +
  tm_shape(rue_rsq_lt0) +
  tm_raster(col.scale = tm_scale_categorical(values = "orange4", labels = expression(Negative~R^2)),
            col.legend = tm_legend(title = "")) +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  borders +
  tm_title("b) RUE")

rpi_rsq_map <- fill +
  tm_shape(rpi_rsq) +
  rsq_raster_params +
  tm_shape(rpi_rsq_lt0) +
  tm_raster(col.scale = tm_scale_categorical(values = "orange4", labels = expression(Negative~R^2)),
            col.legend = tm_legend(title = "", show = FALSE)) +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  borders +
  tm_title("a) RPI")

restrend_rsq_map <- fill +
  tm_shape(restrend_rsq) +
  rsq_raster_params +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  borders +
  tm_title("c) RESTREND")

# rue_rsq_map
# restrend_rsq_map
# rpi_rsq_map

tmap_save(
  rue_rsq_map,
  here("results", "figures", "rue_annual_rsq_map.png"),
  width = 12, height = 12, units = "cm", dpi = 250
)

tmap_save(
  rpi_rsq_map,
  here("results", "figures", "rpi_annual_rsq_map.png"),
  width = 12, height = 12, units = "cm", dpi = 250
)

tmap_save(
  restrend_rsq_map,
  here("results", "figures", "restrend_annual_rsq_map.png"),
  width = 12, height = 12, units = "cm", dpi = 250
)

# Map of best performing strategy

best_method_map <- fill +
  tm_shape(best_rsq) +
  tm_raster(
    col.scale = tm_scale_categorical(
      values = "Set1",
      levels = c(1,2,3),
      labels = c("RUE", "RESTREND", "RPI")
    ),
    col.legend = tm_legend(
      title = "",
      frame = FALSE
    )
  ) +
  tm_title("d) Best method") +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  borders

tmap_save(
  best_method_map,
  here("results", "figures", "best_method_rsq.png"),
  width = 12, height = 12, units = "cm", dpi = 250
)


# # Diff maps
# 
# rpi_rue_rsq_diff_map <- fill +
#   tm_shape(rpi_rue_diff) +
#   tm_raster(
#     col.scale = tm_scale_continuous(
#       limits = c(-1, 1),
#       midpoint = 0,
#       values = "PuOr",
#       outliers.trunc = c(TRUE, TRUE)
#     ),
#     col.legend = tm_legend(
#       title = "RPI vs RUE\nr-squared",
#       reverse = TRUE,
#       width = 10,
#       height = 30
#     )
#   ) +
#   tm_shape(rpi_rsq_lt0) + tm_raster(palette = "grey50", legend.show = FALSE) +
#   tm_shape(rue_rsq_lt0) + tm_raster(palette = "grey50", legend.show = FALSE) +
#   borders
# 
# rpi_restrend_rsq_diff_map <- fill +
#   tm_shape(rpi_restrend_diff) +
#   tm_raster(
#     col.scale = tm_scale_continuous(
#       limits = c(-1, 1),
#       midpoint = 0,
#       values = "PuOr",
#       outliers.trunc = c(TRUE, TRUE)
#     ),
#     col.legend = tm_legend(
#       title = "RPI vs RESTREND\nr-squared",
#       reverse = TRUE,
#       width = 10,
#       height = 30
#     )
#   ) +
#   tm_shape(rpi_rsq_lt0) + tm_raster(palette = "grey50", legend.show = FALSE) +
#   borders
# 
# rpi_rue_rsq_diff_map
# rpi_restrend_rsq_diff_map

```




We can compare the performance of our method with those of others on various
criteria. For mean values across space, the key issues are:
- Explaining a greater share of spatial variation with environmental
covariates (R-squared)
- Reduced overall variance?
- Consistency (lack of sharp boundaries)
- Lack of correlation with MAP or other key vars

For trend, we can assess performance through:
- High R-squared for each pixel
- Low mean absolute error (weighted?) for each pixel
- Consistency in Mann-Kendall trend?
- P-values of Theil-Sen slope or permutation trend?

```{r ppt_correlation}

set.seed(555)
pixel_sample <- terra::spatSample(ppt_all, 1000,
                                  na.rm = TRUE, exhaustive = TRUE,
                                  as.points = TRUE, xy = TRUE)

ppt_sample <- pixel_sample %>%
  as.data.frame(geom = "XY") %>%
  pivot_longer(starts_with("X20"), names_to = "year", values_to = "precipitation") %>%
  mutate(year = as.numeric(substring(year, 2)))

rue_sample <- extract(rue_all, pixel_sample, xy = TRUE) %>%
  pivot_longer(starts_with("gpp_20"), names_to = "year", values_to = "rue") %>%
  mutate(year = as.numeric(substring(year, 5)))
rpi_sample <- extract(rpi_all, pixel_sample) %>%
  pivot_longer(starts_with("rpi_20"), names_to = "year", values_to = "rpi") %>%
  mutate(year = as.numeric(substring(year, 5)))
restrend_sample <- extract(restrend_residuals, pixel_sample) %>%
  pivot_longer(starts_with("restrend_resids_20"), names_to = "year", values_to = "resids") %>%
  mutate(year = as.numeric(substring(year, 17)))

# Combine into a single dataframe
combined_df <- ppt_sample %>%
  left_join(rue_sample) %>%
  left_join(rpi_sample, by = c("ID", "year")) %>%
  left_join(restrend_sample, by = c("ID", "year"))


# Calculate correlations and plot

ppt_cor <- combined_df %>%
  pivot_longer(c(rue, rpi, resids), names_to = "method", values_to = "value") %>%
  group_by(x, y, method) %>%
  summarise(cor = cor(precipitation, value)) %>%
  ungroup()


```

```{r ppt_correlation_map}

raster_to_tidy_df <- function(raster, xy = FALSE) {
  df <- as.data.frame(raster, xy = xy, cells = TRUE)
  
  var_name_year <- names(raster)[1]
  var_name_length <- nchar(var_name_year) - 5
  var_name <- substring(var_name_year, 1, var_name_length)
  
  tidy_df <- df %>%
    pivot_longer(cols = starts_with(var_name),
                 names_to = "year",
                 values_to = var_name) %>%
    mutate(year = as.numeric(substring(year, var_name_length + 2)))
  
  tidy_df
}

ppt_df <- raster_to_tidy_df(ppt_all, TRUE)
rue_df <- raster_to_tidy_df(rue_all)
restrend_df <- raster_to_tidy_df(restrend_residuals)
rpi_df <- raster_to_tidy_df(rpi_all)

combined_df_all <- ppt_df %>%
  left_join(rue_df) %>%
  left_join(restrend_df) %>%
  left_join(rpi_df)

ppt_cor_all <- combined_df_all %>%
  pivot_longer(cols = c(rue, restrend_resids, rpi), names_to = "method") %>%
  group_by(cell, method) %>%
  summarise(ppt_cor = cor(precipitation, value),
            x = first(x),
            y = first(y)) %>%
  ungroup()

# Histogram of correlations

ppt_cor_hist <- ggplot(ppt_cor_all, aes(x = ppt_cor, fill = method)) +
  geom_histogram(position = "identity", bins = 50, colour = "grey20", alpha = 0.5) +
  geom_vline(xintercept = 0, colour = "red", linewidth = 2) +
  theme_bw() +
  labs(x = "Correlation with annual\nprecipitation (pixel-wise)",
       y = "Count",
       fill = "Method")

ggsave(here("results", "figures", "ppt_correlation_hist.png"),
       ppt_cor_hist,
       width = 20, height = 12, units = "cm", dpi = 250)

# Convert back to correlation raster layers

ppt_cor_rast <- ppt_cor_all %>%
  pivot_wider(id_cols = c(cell, x, y),
              values_from = ppt_cor,
              names_from = method) %>%
  select(x, y, rue, restrend, rpi) %>%
  rast(crs = crs(rpi_all), extent = ext(rpi_all))

ppt_cor_rast <- ppt_cor_rast[[c("rue", "restrend_resids", "rpi")]]

# Save maps

ppt_cor_params <- tm_raster(
  col.scale = tm_scale_continuous(
    limits = c(-1, 1),
    midpoint = 0,
    values = "PuOr"
  ),
  col.legend = tm_legend(
    title = "Correlation",
    frame = FALSE,
    reverse = TRUE
  )
)

map_layout <- tm_layout(
  asp = 3/4,
  frame = FALSE
)

rpi_ppt_cor <- fill +
  tm_shape(ppt_cor_rast$rpi) +
  ppt_cor_params +
  map_layout +
  borders +
  tm_title("a) RPI")

rue_ppt_cor <- fill +
  tm_shape(ppt_cor_rast$rue) +
  ppt_cor_params +
  map_layout +
  borders +
  tm_title("b) RUE")

restrend_ppt_cor <- fill +
  tm_shape(ppt_cor_rast$restrend_resids) +
  ppt_cor_params +
  map_layout +
  borders +
  tm_title("c) RESTREND")

best_cor <- which.min(abs(ppt_cor_rast))

best_ppt_cor <- fill +
  tm_shape(best_cor) +
  tm_raster(
    col.scale = tm_scale_categorical(
      values = "Set2",
      levels = c(1,2,3),
      labels = c("RUE", "RESTREND", "RPI")
    ),
    col.legend = tm_legend(
      title = "",
      frame = FALSE
    )
  ) +
  map_layout +
  borders +
  tm_title("d) Best")

tmap_save(
  rpi_ppt_cor,
  here("results", "figures", "rpi_ppt_cor.png"),
  width = 12, height = 12, units = "cm", dpi = 250
)

tmap_save(
  rue_ppt_cor,
  here("results", "figures", "rue_ppt_cor.png"),
  width = 12, height = 12, units = "cm", dpi = 250
)

tmap_save(
  restrend_ppt_cor,
  here("results", "figures", "restrend_ppt_cor.png"),
  width = 12, height = 12, units = "cm", dpi = 250
)

tmap_save(
  best_ppt_cor,
  here("results", "figures", "best_ppt_cor.png"),
  width = 12, height = 12, units = "cm", dpi = 250
)

```

```{r fit_histograms}

indices <- c("RUE", "RESTREND", "RPI")

# R-squared
combined_rsq <- c(rue_rsq, restrend_rsq, rpi_rsq)

names(combined_rsq) <- indices

rsq_df <- as.data.frame(combined_rsq, na.rm = T)

rsq_hist <- rsq_df %>%
  pivot_longer(cols = all_of(indices),
               names_to = "method", values_to = "rsq") %>%
  mutate(method = ordered(method, levels = c("RUE", "RESTREND", "RPI"))) %>%
  ggplot(aes(x = rsq, fill = method)) +
  geom_density(position = "identity", alpha = 0.5, colour = "grey30") +
  geom_vline(xintercept = 0) +
  theme_bw() +
  scale_fill_brewer(palette = "Set2") +
  xlim(-2, 1) +
  labs(x = expression("Pixel-wise"~R^2),
       y = "Density",
       fill = "Method")

# MAE

combined_mae <- c(mae_rue, mae_restrend, mae_rpi_anom)
names(combined_mae) <- indices

mae_df <- as.data.frame(combined_mae, na.rm = T)

mae_hist <- mae_df %>%
  pivot_longer(cols = all_of(indices),
               names_to = "method", values_to = "mae") %>%
  mutate(method = ordered(method, levels = c("RUE", "RESTREND", "RPI"))) %>%
  ggplot(aes(x = mae, fill = method)) +
  geom_density(position = "identity", alpha = 0.5, colour = "grey30") +
  # geom_vline(xintercept = 0) +
  theme_bw() +
  scale_fill_brewer(palette = "Set2") +
  xlim(0, 1500) +
  labs(x = "Mean absolute error",
       y = "Density",
       fill = "Method")


# Save plots

ggsave(
  here("results", "figures", "rsq_hist.png"),
  rsq_hist,
  width = 16, height = 12, units = "cm", dpi = 250
)

ggsave(
  here("results", "figures", "mae_hist.png"),
  mae_hist,
  width = 16, height = 12, units = "cm", dpi = 250
)

```



Now to actually plot the trends of the three time series methods:

```{r ts_trend}

# Basemaps

fill <- tm_shape(ke_tz) +
  tm_fill("grey95")

borders <- tm_shape(ke_tz) +
  tm_borders()

map_layout <- tm_layout(
  asp = 3/4,
  frame = FALSE
)

# Theil-Sen maps

create_ts_trend_map <- function(trend_raster, palette) {
  quantiles <- quantile(values(trend_raster), c(0.01, 0.99), na.rm = T)
  
  tm_shape(trend_raster) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = quantiles,
      midpoint = 0,
      values = palette,
      outliers.trunc = c(TRUE, TRUE)
    ),
    col.legend = tm_legend(
      title = "Slope",
      reverse = TRUE,
      frame = TRUE
    )
  )
}

rue_ts_map <- fill +
  create_ts_trend_map(rue_ts_trend[[1]], "PuOr") +
  borders +
  map_layout +
  tm_title("b) RUE")

rpi_ts_map <- fill +
  create_ts_trend_map(rpi_ts_trend[[1]], "PuOr") +
  borders +
  map_layout +
  tm_title("a) RPI")

restrend_ts_map <- fill +
  create_ts_trend_map(restrend[[1]], "PuOr") +
  borders +
  map_layout +
  tm_title("c) RESTREND")


tmap_save(rue_ts_map,
          here("results", "figures", "rue_ts_map.png"),
          width = 12, height = 12, units = "cm", dpi = 250)
tmap_save(rpi_ts_map,
          here("results", "figures", "rpi_ts_map.png"),
          width = 12, height = 12, units = "cm", dpi = 250)
tmap_save(restrend_ts_map,
          here("results", "figures", "restrend_ts_map.png"),
          width = 12, height = 12, units = "cm", dpi = 250)



# Kendall Tau maps

kendall_tau_params <- tm_raster(
  col.scale = tm_scale_continuous(
    limits = c(-0.5, 0.5),
    midpoint = 0,
    values = "PuOr",
    outliers.trunc = c(TRUE, TRUE)
  ),
  col.legend = tm_legend(
    reverse = TRUE,
    title = "Kendall tau",
    frame = FALSE
  )
)

rue_kendall_map <- fill +
  tm_shape(rue_kendall[[1]]) +
  kendall_tau_params +
  borders +
  map_layout +
  tm_title("b) RUE")

rpi_kendall_map <- fill +
  tm_shape(rpi_kendall[[1]]) +
  kendall_tau_params +
  borders +
  map_layout +
  tm_title("a) RPI")

restrend_kendall_map <- fill +
  tm_shape(restrend_kendall[[1]]) +
  kendall_tau_params +
  borders +
  map_layout +
  tm_title("c) RESTREND")


tmap_save(rue_kendall_map,
          here("results", "figures", "rue_kendall_map.png"),
          width = 12, height = 12, units = "cm", dpi = 250)
tmap_save(rpi_kendall_map,
          here("results", "figures", "rpi_kendall_map.png"),
          width = 12, height = 12, units = "cm", dpi = 250)
tmap_save(restrend_kendall_map,
          here("results", "figures", "restrend_kendall_map.png"),
          width = 12, height = 12, units = "cm", dpi = 250)

# Pixels with opposite trends in RPI and RESTREND




```




```{r trend_scatter_plots}

options(scipen = 999)

set.seed(121)
rpi_trend_sample <- spatSample(rpi_ts_trend$slope, 4000, na.rm = TRUE, exhaustive = TRUE, as.points = TRUE)

rue_trend_sample <- extract(rue_ts_trend$rue_slope, rpi_trend_sample)
restrend_sample <- extract(restrend$resid_slope, rpi_trend_sample)
mean_ppt_sample <- ppt_all %>%
  mean() %>%
  extract(rpi_trend_sample)

trend_sample_all <- rpi_trend_sample %>%
  as.data.frame(geom = "XY") %>%
  bind_cols(rue_trend_sample, restrend_sample, mean_ppt_sample) %>%
  select(-starts_with("ID")) %>%
  rename(precipitation = mean,
         rpi_slope = slope,
         restrend_slope = resid_slope) %>%
  mutate(ppt_bin = cut_width(precipitation, 200, boundary = 0))

bins <- levels(trend_sample_all$ppt_bin)

trend_sample_subset <- filter(trend_sample_all, ppt_bin %in% bins[1:6])

rpi_rue_plot <- ggplot(trend_sample_subset, aes(x = rpi_slope, y = rue_slope)) +
  geom_point(size = 0.5) +
  xlim(-0.03, 0.03) +
  ylim(-0.2, 0.2) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  facet_wrap(~ppt_bin) +
  theme_bw() +
  labs(x = "RPI", y = "RUE")

rpi_restrend_plot <- ggplot(trend_sample_subset, aes(x = rpi_slope, y = restrend_slope)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  xlim(-0.03, 0.03) +
  # ylim(-0.2, 0.2) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  facet_wrap(~ppt_bin) +
  theme_bw() +
  labs(x = "RPI", y = "RESTREND")

rpi_rue_plot
rpi_restrend_plot

# Correlations within rainfall buckets (use whole dataset)

slope_raster <- c(rpi_ts_trend$slope, rue_ts_trend$rue_slope, restrend$resid_slope, mean(ppt_all))
names(slope_raster) <- c("rpi_slope", "rue_slope", "restrend_slope", "mean_ppt")

slope_df <- as.data.frame(slope_raster, cells = TRUE, na.rm = TRUE)

slope_df_binned <- slope_df %>%
  mutate(ppt_bin = cut_width(mean_ppt, width = 200, boundary = 0)) %>%
  group_by(ppt_bin) %>%
  filter(n() >= 1000) %>%
  ungroup()

rpi_restrend_cor <-  slope_df_binned %>%
  group_by(ppt_bin) %>%
  summarise(rpi_rue_cor = cor(rpi_slope, rue_slope),
            rpi_restrend_cor = cor(rpi_slope, restrend_slope))

ppt_bin_sample <- slope_df_binned %>%
  group_by(ppt_bin) %>%
  slice_sample(n = 1000)

rpi_rue_plot <- ggplot(ppt_bin_sample, aes(x = rpi_slope, y = rue_slope)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, lwd = 0.8, colour = "grey60") +
  xlim(-0.03, 0.03) +
  ylim(-0.2, 0.2) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  facet_wrap(~ppt_bin) +
  theme_bw() +
  labs(x = "RPI", y = "RUE")

rpi_restrend_plot <- ggplot(ppt_bin_sample, aes(x = rpi_slope, y = restrend_slope)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, lwd = 0.8, colour = "grey60") +
  xlim(-0.05, 0.05) +
  ylim(-150, 150) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  facet_wrap(~ppt_bin) +
  theme_bw() +
  labs(x = "RPI", y = "RESTREND")

```



```{r facet_maps_rpi}

names(rpi_all) <- years

rpi_annual_map <- fill +
  tm_shape(rpi_all[[2:19]]) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(0.2, 1.2),
      midpoint = 0.7,
      values = "viridis",
      outliers.trunc = c(TRUE, TRUE)
    ),
    col.legend = tm_legend(show = FALSE)
  ) +
  borders +
  tm_layout(asp = 3/4, frame = FALSE,
            panel.label.bg.color = "white",
            panel.label.size = 10) +
  tm_facets_wrap(nrows = 3, ncols = 6)

rpi_map_legend <- tm_shape(rpi_all[[2]]) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(0.2, 1.2),
      midpoint = 0.7,
      values = "viridis",
      values.range = c(0,1),
      outliers.trunc = c(TRUE, TRUE)
    ),
    col.legend = tm_legend(
      title = "RPI",
      reverse = TRUE,
      frame = FALSE
    ),
  ) +
  tm_layout(legend.only = TRUE)

tmap_save(
  rpi_annual_map,
  here("results", "figures", "rpi_annual_map.png"),
  width = 30, height = 20, units = "cm", dpi = 250
)

tmap_save(
  rpi_map_legend,
  here("results", "figures", "rpi_map_legend.png"),
  width = 3, height = 6, units = "cm", dpi = 250
)

```


```{r rue_new}

gpp_ppt <- c(actual_gpp_all, ppt_all)

rue_lm <- app(gpp_ppt, function(x) {
  if (any(is.na(x))) {
    c(NA, NA)
  } else {
    gpp_x <- x[1:19]
    ppt_x <- x[20:38]
    
    mod <- lm(gpp_x ~ 0 + ppt_x)
    
    slope <- coef(mod)[1]
    
    rsq <- broom::glance(mod)$r.squared
    
    c(slope, rsq)
  }
})

writeRaster(rue_lm,
            here("data", "processed", "raster", "rue", "rue_lm.tif"))

```



```{r rue_new_test}

rue_lm <- rast(here("data", "processed", "raster", "rue", "rue_lm.tif"))

plot(rue_lm)

rue_resids <- actual_gpp_all - (rue_lm * ppt_all)




```



```{r rpi_autocorrelation}

rpi_detrended <- app(rpi_all, function(x) {
  
  if(any(is.na(x))) {
    rep(NA, 19)
  } else {
    years <- 1:19
    
    trend <- lm(x ~ years)
    
    residuals(trend)
    
  }
})

calc_ac1 <- function(x) {
  if(any(is.na(x))) {
    NA
  } else {
    ac_all <- acf(x, lag.max = 2, plot = FALSE)
    
    ac_all$acf[2]
  }
}

rpi_ac1 <- app(rpi_all, calc_ac1)
rpi_ac1_detrended <- app(rpi_detrended, calc_ac1)

# RESTREND??

restrend_detrended <- app(restrend_residuals, function(x) {
  
  if(any(is.na(x))) {
    rep(NA, 19)
  } else {
    years <- 1:19
    
    trend <- lm(x ~ years)
    
    residuals(trend)
    
  }
})

restrend_ac1 <- app(restrend_residuals, calc_ac1)
restrend_ac1_detrended <- app(restrend_detrended, calc_ac1)

# GPP only

gpp_ac1 <- app(actual_gpp_all, calc_ac1)

# RPI vs previous year's precipitation
rpi_df <- as.data.frame(rpi_all, cells = TRUE, na.rm = TRUE)
ppt_df <- as.data.frame(ppt_all, cells = TRUE, na.rm = TRUE)

rpi_ppt_df <- left_join(rpi_df, ppt_df)

rpi_ppt_pivot <- rpi_ppt_df %>%
  pivot_longer(cols = starts_with(c("rpi", "precipitation"))) %>%
  separate_wider_delim(cols = "name", delim = "_", names = c("index", "year")) %>%
  mutate(year = as.numeric(year)) %>%
  pivot_wider(id_cols = c(cell, year),
              names_from = "index", values_from = "value")

rpi_ppt_ac1 <- rpi_ppt_pivot %>%
  group_by(cell) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(ppt_lag = lag(precipitation)) %>%
  filter(year > 2001) %>%
  summarise(cor = cor(rpi, ppt_lag))

median(rpi_ppt_ac1$cor)

```

```{r palette_tester_array}

# R-G palette reflects positive/negative trend
# B reflects current value
n <- 5

r <- matrix(rep(1:n, n), nrow = n, byrow = FALSE)
g <- n + 1 - r
b <- matrix(rep(1:n, n), nrow = n, byrow = TRUE)

a <- array(c(r,g,b), dim = c(n,n,3))
R <- rast(a)

labels <- c("R", "G", "B")
names(R) <- labels

tm_shape(R) + 
  tm_rgb(
    col = tm_mv(labels),
    col.scale = tm_scale_rgb(maxValue = n)
  )


# Not quite. There's too much overlap.
# How about reducing the overlap between R and G?

n2 <- 2 * n - 1

r2 <- matrix(
  c(
    rep(0, n * (n - 1)),
    rep(1:n, each = n)
  ),
  nrow = n2,
  byrow = TRUE
)
g2 <- matrix(
  c(
    rep(n:1, each = n),
    rep(0, n * (n - 1))
  ),
  nrow = n2,
  byrow = TRUE
)
b2 <- matrix(
  rep(1:n, each = n2),
  nrow = n2,
  byrow = FALSE
)

a2 <- array(c(r2, g2, b2), dim = c(n2, n, 3))
R2 <- rast(a2)

names(R2) <- labels

tm_shape(R2) + 
  tm_rgb(
    col = tm_mv(labels),
    col.scale = tm_scale_rgb(maxValue = n)
  )

# I don't think this is clear enough. There's too much colour variation
# and it's too hard to distinguish productivity levels in any given part
# of the spectrum. I should use HSL as it's more akin to how humans perceive
# colour.

# For HSL, I need to use hue as the metric for trend, since saturation and
# luminance don't have anything like a midpoint.


```


