---
title: "Results"
output: html_notebook
author: Guy Lomax
date: 2024-01-12
editor_options: 
  chunk_output_type: console
---

This notebook compares the results of the RPI methodology with those of RUE,
RESTREND and LGS.

```{r setup}

library(tidyverse)
library(terra)
library(sf)
library(here)
library(tmap)

```


```{r load}

# Country polygons
ke_tz <- st_read(here("data", "raw", "vector", "kenya_tanzania.geojson"))

# RPI mean and trend
rpi_mean <- rast(here("data", "processed", "raster", "rpi", "rpi_mean.tif"))
# rpi_ts_trend <- rast(here("data", "processed", "raster", "rpi", "rf_rpi_ts_trend.tif"))

# RUE mean and trend
rue_mean <- rast(here("data", "processed", "raster", "rue", "rue_mean.tif"))
# rue_ts_trend <- rast(here("data", "processed", "raster", "rue", "rue_trend.tif"))

# RESTREND trend
restrend <- rast(here("data", "processed", "raster", "restrend", "restrend_results.tif"))

# LGS mean
lgs_100 <- rast(here("data", "processed", "raster", "lgs", "lgs_rast_100.tif"))
lgs_50 <- rast(here("data", "processed", "raster", "lgs", "lgs_rast_50.tif"))
lgs_20 <- rast(here("data", "processed", "raster", "lgs", "lgs_rast_20.tif"))

# Key covariates for comparison/exploration of results
dynamic_covariate_paths <- Sys.glob(here("data", "raw", "raster", "covariateMaps", "dynamicVars*.tif"))
ppt_mean <- dynamic_covariate_paths %>%
  map(function(path) rast(path) %>% subset("precipitation")) %>%
  rast() %>%
  mean()
t_mean <- dynamic_covariate_paths %>%
  map(function(path) rast(path) %>% subset("tMean")) %>%
  rast() %>%
  mean()
tree_cover <- rast(here("data", "raw", "raster", "covariateMaps", "staticVars.tif")) %>%
  subset("wriTCFrac") %>%
  crop(ppt_mean)

covariates <- c(ppt_mean, t_mean, tree_cover)
names(covariates) <- c("ppt_mean", "t_mean", "tree_cover")
  

```


First, we can simply plot comparative maps of the results from the different
methods, starting with those that allow spatial comparison of mean degradation
or productivity.

```{r spatial_comparison_maps}

# Basemap elements
fill <- tm_shape(ke_tz) +
  tm_fill("grey95")

borders <- tm_shape(ke_tz) +
  tm_borders() +
  tm_layout(frame = FALSE)

# Raw maps of mean values
rpi_mean_map <- fill +
  tm_shape(rpi_mean) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(0.3, 1),
      midpoint = 0.65,
      values = "RdBu",
      outliers.trunc = c(T,T)
    ),
    col.legend = tm_legend(
      title = "Mean RPI",
      reverse = TRUE,
      width = 10,
      height = 30)
  ) +
  borders

rue_mean_map <- fill +
  tm_shape(rue_mean) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(0, 7),
      midpoint = 3,
      values = "RdBu",
      outliers.trunc = c(T,T)
    ),
    col.legend = tm_legend(
      title = "Mean RUE",
      reverse = TRUE,
      width = 10,
      height = 30)
  ) +
  borders

lgs_mean_map <- fill +
  tm_shape(lgs_100$lgs_ratio) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(0.4, 1.2),
      midpoint = 0.8,
      values = "RdBu",
      outliers.trunc = c(T,T)
    ),
    col.legend = tm_legend(
      title = "LGS",
      reverse = TRUE,
      width = 10,
      height = 30)
  ) +
  borders

rpi_mean_map
rue_mean_map
lgs_mean_map

## Comparative map of RPI vs. LGS (scaled relative to median and 0.1/0.9 quantiles)

# Calculate median and quartiles for each index
rpi_quantiles <- quantile(values(rpi_mean), c(0.5, 0.25, 0.75), na.rm = TRUE)
rue_quantiles <- quantile(values(rue_mean), c(0.5, 0.25, 0.75), na.rm = TRUE)
lgs_quantiles <- quantile(values(lgs_100$lgs_ratio), c(0.5, 0.25, 0.75), na.rm = TRUE)

# Centre values around median and scale by interquartile range
rpi_mean_scaled <- (rpi_mean - rpi_quantiles[1]) / (rpi_quantiles[3] - rpi_quantiles[2])
rue_mean_scaled <- (rue_mean - rue_quantiles[1]) / (rue_quantiles[3] - rue_quantiles[2])
lgs_mean_scaled <- (lgs_100$lgs_ratio - lgs_quantiles[1]) / (lgs_quantiles[3] - lgs_quantiles[2])

# Calculate difference maps between RPI and other indices

rpi_rue_diff <- rpi_mean_scaled - rue_mean_scaled
rpi_lgs_diff <- rpi_mean_scaled - lgs_mean_scaled

rpi_rue_diff_map <- fill +
  tm_shape(rpi_rue_diff) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(-2, 2),
      midpoint = 0,
      values = "RdBu",
      outliers.trunc = c(T,T)
    ),
    
    col.legend = tm_legend(
      title = "RPI vs. RUE difference",
      reverse = TRUE,
      width = 10,
      height = 30
    )
  ) +
  borders

rpi_lgs_diff_map <- fill +
  tm_shape(rpi_lgs_diff) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(-2, 2),
      midpoint = 0,
      values = "RdBu",
      outliers.trunc = c(T,T)
    ),
    
    col.legend = tm_legend(
      title = "RPI vs. LGS difference",
      reverse = TRUE,
      width = 10,
      height = 30
    )
  ) +
  borders

rpi_rue_diff_map
rpi_lgs_diff_map


```


A more intuitive map might be to plot the quantiles for each dataset, i.e.,
which quantile does each fit into in terms of overall productivity. That is a
scale-free metric.

```{r comparison_maps_quantiles}

rast_as_quantiles <- function(x, quantile_list) {
  
  x_values <- values(x)
  quantiles <- quantile(x_values, quantile_list, na.rm = TRUE)
  
  quantile_values <- findInterval(x_values, quantiles)
  
  out <- quantile_values %>%
    matrix(nrow = nrow(x), ncol = ncol(x), byrow = TRUE) %>%
    rast(extent = ext(x), crs = crs(x))
  
  out
}

quantiles <- seq(0, 1, 0.01)

rpi_quantile_rast <- rast_as_quantiles(rpi_mean, quantiles)
rue_quantile_rast <- rast_as_quantiles(rue_mean, quantiles)
lgs_quantile_rast <- rast_as_quantiles(lgs_100$lgs_ratio, quantiles)

plot(rpi_quantile_rast)
plot(rue_quantile_rast)
plot(lgs_quantile_rast)

# Plot quantile difference maps

# Basemap elements
fill <- tm_shape(ke_tz) +
  tm_fill("grey95")

borders <- tm_shape(ke_tz) +
  tm_borders() +
  tm_layout(frame = FALSE)

rpi_rue_diff_map_q <- fill +
  tm_shape(rpi_quantile_rast - rue_quantile_rast) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(-60, 60),
      midpoint = 0,
      values = "PuOr",
      outliers.trunc = c(TRUE, TRUE)
    ),
    col.legend = tm_legend(
      reverse = TRUE,
      title = "RPI - RUE\n(quantiles)",
      width = 10,
      height = 30
    )
  ) +
  borders

rpi_lgs_diff_map_q <- fill +
  tm_shape(rpi_quantile_rast - lgs_quantile_rast) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(-60, 60),
      midpoint = 0,
      values = "PuOr",
      outliers.trunc = c(TRUE, TRUE)
    ),
    col.legend = tm_legend(
      reverse = TRUE,
      title = "RPI - LGS\n(quantiles)",
      width = 10,
      height = 30
    )
  ) +
  borders

rpi_rue_diff_map_q
rpi_lgs_diff_map_q


```


We can also plot a sample of points from each method to explore the
quantitative relationship between them.

```{r spatial_comparison_plots}

set.seed(999)
mean_rpi_sample <- spatSample(rpi_mean, 2000, as.points = TRUE,
                              na.rm = TRUE, exhaustive = TRUE)

mean_rue_sample <- extract(rue_mean, mean_rpi_sample)

mean_lgs_sample <- extract(lgs_100$lgs_ratio, mean_rpi_sample)

covariate_sample <- extract(covariates, mean_rpi_sample)

spatial_comparison <- mean_rpi_sample %>%
  st_as_sf() %>%
  rename(mean_rpi = mean) %>%
  bind_cols(mean_rue_sample) %>%
  rename(mean_rue = mean) %>%
  bind_cols(mean_lgs_sample) %>%
  bind_cols(covariate_sample) %>%
  select(-starts_with("ID"))


ggplot(spatial_comparison, aes(x = mean_rpi, y = mean_rue)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()

ggplot(spatial_comparison, aes(x = mean_rpi, y = lgs_ratio)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()

spatial_comparison_valid <- drop_na(spatial_comparison)

cor(spatial_comparison_valid$mean_rpi, spatial_comparison_valid$mean_rue)
cor(spatial_comparison_valid$mean_rpi, spatial_comparison_valid$lgs_ratio)

# Results relative to mean precipitation
spatial_comparison %>%
  pivot_longer(1:3, names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = ppt_mean, y = value)) +
  geom_point(size = 0.25) +
  geom_smooth(method = "lm", se = F) +
  facet_wrap(~metric, scales = "free") +
  theme_bw() +
  labs(x = "Mean annual\nprecipitation (mm)",
       y = "Metric")

spatial_comparison %>%
  st_drop_geometry %>%
  pivot_longer(1:3, names_to = "metric", values_to = "value") %>%
  group_by(metric) %>%
  drop_na() %>%
  summarise(cor = cor(ppt_mean, value))


```


We can compare the performance of our method with those of others on various
criteria. For mean values across space, the key issues are:
- Explaining a greater share of spatial variation with environmental
covariates (R-squared)
- Reduced overall variance?
- Consistency (lack of sharp boundaries)
- Lack of correlation with MAP or other key vars

For trend, we can assess performance through:
- High R-squared for each pixel
- Low mean absolute error (weighted?) for each pixel
- Consistency in Mann-Kendall trend?
- P-values of Theil-Sen slope or permutation trend?
