---
title: "Results"
output: html_notebook
author: Guy Lomax
date: 2024-01-12
editor_options: 
  chunk_output_type: console
---

This notebook compares the results of the RPI methodology with those of RUE,
RESTREND and LGS at identifying persistent differences in multi-annual mean
productivity.

```{r setup}

library(tidyverse)
library(terra)
library(sf)
library(here)
library(tmap)

```


```{r load}

# Country polygons
ke_tz <- st_read(here("data", "raw", "vector", "kenya_tanzania.geojson"))

# Actual GPP mean
rpi_paths <- Sys.glob(here("data", "processed", "raster", "rpi", "rf_rpi_100_20*.tif"))

potential_gpp_mean <- rpi_paths %>%
  map(function(path) rast(path) %>% subset("quantile_pred")) %>%
  rast() %>%
  mean()

actual_gpp_mean <- rpi_paths %>%
  map(function(path) rast(path) %>% subset("GPP")) %>%
  rast() %>%
  mean()

predicted_gpp_rpi <- rpi_paths %>%
  map(function(path) rast(path) %>% subset("mean_pred")) %>%
  rast() %>%
  mean()

# RPI mean and trend
rpi_mean <- rast(here("data", "processed", "raster", "rpi", "rpi_mean.tif"))

# RUE mean and trend
rue_mean <- rast(here("data", "processed", "raster", "rue", "rue_mean.tif"))

# LGS mean

lgs_k <- c(20, 50, 100, 200, 500, 1000)
lgs_paths <- Sys.glob(here("data", "processed", "raster", "lgs", "lgs_subset_*.tif"))

lgs_layers <- lgs_paths %>%
  map(rast)
names(lgs_layers) <- lgs_k

lgs_50 <- lgs_layers[["50"]]
lgs_100 <- lgs_layers[["100"]]
lgs_200 <- lgs_layers[["200"]]
lgs_1000 <- lgs_layers[["1000"]]

# Key covariates for comparison/exploration of results
dynamic_covariate_paths <- Sys.glob(here("data", "raw", "raster", "covariateMaps", "dynamicVars*.tif"))

years <- 2000:2018
ppt_all <- dynamic_covariate_paths %>%
  map(function(path) rast(path) %>% subset("precipitation")) %>%
  rast()
t_mean_all <- dynamic_covariate_paths %>%
  map(function(path) rast(path) %>% subset("tMean")) %>%
  rast()
tree_cover <- rast(here("data", "raw", "raster", "covariateMaps", "staticVars.tif")) %>%
  subset("wriTCFrac") %>%
  crop(ppt_all)

names(ppt_all) <- paste0("ppt_", years)
names(t_mean_all) <- paste0("tMean_", years)
names(tree_cover) <- "tree_cover"

ppt_mean <- mean(ppt_all)
t_mean <- mean(t_mean_all)
covariates <- c(ppt_mean, t_mean, tree_cover)
names(covariates) <- c("ppt_mean", "t_mean", "tree_cover")


```


We can estimate the usefulness of each method in comparing spatial patterns
by quantifying how much of the spatial variation in GPP it is able to explain
using environmental variation. For RUE this is extremely crude, but for the
others it is more meaningful.

```{r spatial_variance_explained}

mean_gpp_value <- mean(values(actual_gpp_mean), na.rm = T)

tss_gpp <- (actual_gpp_mean - mean_gpp_value) ^ 2 %>%
  values() %>%
  sum(na.rm = TRUE)

# RUE

median_rue <- median(values(rue_mean), na.rm = T)
mean_rue <- mean(values(rue_mean), na.rm = T)

predicted_gpp_rue <- median_rue * mean(ppt_all)

abs_error_rue <- abs(actual_gpp_mean - predicted_gpp_rue)

mae_rue <- mean(values(abs_error_rue), na.rm = TRUE)

rss_gpp_rue <- (actual_gpp_mean - predicted_gpp_rue) ^ 2 %>%
  values() %>%
  sum(na.rm = TRUE)

rsq_rue <- 1 - (rss_gpp_rue / tss_gpp)

# LGS

gt_weight <- 5
lte_weight <- 5 / 9

error_lgs <- actual_gpp_mean - lgs_1000$potential

weighted_abs_error_lgs <- ifel(error_lgs > 0,
                               error_lgs * gt_weight,
                               error_lgs * lte_weight) %>%
  abs()

maew_lgs <- mean(values(weighted_abs_error_lgs), na.rm = T)

mae_lgs <- mean(values(abs(actual_gpp_mean - lgs_1000$mean)), na.rm = T)


rss_gpp_lgs <- (actual_gpp_mean - lgs_1000$mean) ^ 2 %>%
  values() %>%
  sum(na.rm = T)

rsq_lgs <- 1 - (rss_gpp_lgs / tss_gpp)

# RPI

error_rpi <- actual_gpp_mean - potential_gpp_mean

weighted_abs_error_rpi <- ifel(error_rpi > 0,
                               error_rpi * gt_weight,
                               error_rpi * lte_weight) %>%
  abs()

maew_rpi <- mean(values(weighted_abs_error_rpi), na.rm = T)

mae_rpi <- mean(values(abs(actual_gpp_mean - predicted_gpp_rpi)), na.rm = T)

rss_gpp_rpi <- (actual_gpp_mean - predicted_gpp_rpi) ^ 2 %>%
  values() %>%
  sum(na.rm = TRUE)

rsq_rpi <- 1 - (rss_gpp_rpi / tss_gpp)

```





First, we can simply plot comparative maps of the results from the different
methods, starting with those that allow spatial comparison of mean degradation
or productivity.

```{r spatial_comparison_maps}

# Basemap elements
fill <- tm_shape(ke_tz) +
  tm_fill("grey95")

borders <- tm_shape(ke_tz) +
  tm_borders() +
  tm_layout(frame = FALSE)

# Raw maps of mean values
rpi_mean_map <- fill +
  tm_shape(rpi_mean) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(0.3, 1),
      values = "YlGnBu",
      outliers.trunc = c(T,T)
    ),
    col.legend = tm_legend(
      title = "Mean RPI",
      reverse = TRUE,
      frame = FALSE)
  ) +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  tm_title("a) RPI") +
  borders

rue_mean_map <- fill +
  tm_shape(rue_mean) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(1, 7),
      values = "YlGnBu",
      outliers.trunc = c(T,T)
    ),
    col.legend = tm_legend(
      title = "Mean RUE",
      reverse = TRUE,
      frame = FALSE)
  ) +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  tm_title("b) RUE") +
  borders

lgs_100_mean_map <- fill +
  tm_shape(lgs_100$lgs_ratio) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(0.2, 1.4),
      values = "YlGnBu",
      outliers.trunc = c(T,T)
    ),
    col.legend = tm_legend(
      title = "LGS",
      reverse = TRUE,
      frame = FALSE)
  ) +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  tm_title("c) LGS, k = 100") +
  borders

lgs_200_mean_map <- fill +
  tm_shape(lgs_200$lgs_ratio) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(0.2, 1.4),
      values = "YlGnBu",
      outliers.trunc = c(T,T)
    ),
    col.legend = tm_legend(
      title = "LGS",
      reverse = TRUE,
      frame = FALSE)
  ) +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  tm_title("d) LGS, k = 200") +
  borders

lgs_1000_mean_map <- fill +
  tm_shape(lgs_1000$lgs_ratio) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(0.2, 1.4),
      values = "YlGnBu",
      outliers.trunc = c(T,T)
    ),
    col.legend = tm_legend(
      title = "LGS",
      reverse = TRUE,
      frame = FALSE)
  ) +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  tm_title("d) LGS, k = 1000") +
  borders

# Save maps
tmap_save(rue_mean_map,
          here("results", "figures", "rue_mean.png"),
          width = 12, height = 12, units = "cm", dpi = 250)

tmap_save(rpi_mean_map,
          here("results", "figures", "rpi_mean.png"),
          width = 12, height = 12, units = "cm", dpi = 250)

tmap_save(lgs_100_mean_map,
          here("results", "figures", "lgs_100_mean.png"),
          width = 12, height = 12, units = "cm", dpi = 250)

tmap_save(lgs_200_mean_map,
          here("results", "figures", "lgs_200_mean.png"),
          width = 12, height = 12, units = "cm", dpi = 250)

tmap_save(lgs_1000_mean_map,
          here("results", "figures", "lgs_1000_mean.png"),
          width = 12, height = 12, units = "cm", dpi = 250)

# Potential GPP maps for RPI and LGS
rpi_potential_map <- fill +
  tm_shape(potential_gpp_mean) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(750, 5250),
      midpoint = 3000,
      values = "YlGn",
      outliers.trunc = c(TRUE, TRUE)
    ),
    col.legend = tm_legend(
      reverse = TRUE,
      title = expression(
        atop("Potential GPP",
             kg~C~m^-2~y^-1)
      ),
      frame = FALSE
    )
  ) +
  borders +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  tm_title("RPI")

lgs_100_potential_map <- fill +
  tm_shape(lgs_100$potential) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(750, 5250),
      midpoint = 3000,
      values = "YlGn",
      outliers.trunc = c(TRUE, TRUE)
    ),
    col.legend = tm_legend(
      reverse = TRUE,
      title = expression(
        atop("Potential GPP",
             kg~C~m^-2~y^-1)
      ),
      frame = FALSE
    )
  ) +
  borders +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  tm_title("LGS, k = 100")

lgs_1000_potential_map <- fill +
  tm_shape(lgs_1000$potential) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(750, 5250),
      midpoint = 3000,
      values = "YlGn",
      outliers.trunc = c(TRUE, TRUE)
    ),
    col.legend = tm_legend(
      reverse = TRUE,
      title = expression(
        atop("Potential GPP",
             kg~C~m^-2~y^-1)
      ),
      frame = FALSE
    )
  ) +
  borders +
  tm_layout(asp = 3/4,
            frame = FALSE) +
  tm_title("LGS, k = 1000")


tmap_save(rpi_potential_map,
          here("results", "figures", "rpi_potential_map.png"),
          width = 12, height = 12, units = "cm", dpi = 250)

tmap_save(lgs_100_potential_map,
          here("results", "figures", "lgs_100_potential_map.png"),
          width = 12, height = 12, units = "cm", dpi = 250)

tmap_save(lgs_1000_potential_map,
          here("results", "figures", "lgs_1000_potential_map.png"),
          width = 12, height = 12, units = "cm", dpi = 250)

rpi_mean_map
rue_mean_map
lgs_mean_map

# ## Comparative map of RPI vs. LGS (scaled relative to median and 0.1/0.9 quantiles)
# 
# # Calculate median and quartiles for each index
# rpi_quantiles <- quantile(values(rpi_mean), c(0.5, 0.25, 0.75), na.rm = TRUE)
# rue_quantiles <- quantile(values(rue_mean), c(0.5, 0.25, 0.75), na.rm = TRUE)
# lgs_quantiles <- quantile(values(lgs_100$lgs_ratio), c(0.5, 0.25, 0.75), na.rm = TRUE)
# 
# # Centre values around median and scale by interquartile range
# rpi_mean_scaled <- (rpi_mean - rpi_quantiles[1]) / (rpi_quantiles[3] - rpi_quantiles[2])
# rue_mean_scaled <- (rue_mean - rue_quantiles[1]) / (rue_quantiles[3] - rue_quantiles[2])
# lgs_mean_scaled <- (lgs_100$lgs_ratio - lgs_quantiles[1]) / (lgs_quantiles[3] - lgs_quantiles[2])
# 
# # Calculate difference maps between RPI and other indices
# 
# rpi_rue_diff <- rpi_mean_scaled - rue_mean_scaled
# rpi_lgs_diff <- rpi_mean_scaled - lgs_mean_scaled
# 
# rpi_rue_diff_map <- fill +
#   tm_shape(rpi_rue_diff) +
#   tm_raster(
#     col.scale = tm_scale_continuous(
#       limits = c(-2, 2),
#       midpoint = 0,
#       values = "RdBu",
#       outliers.trunc = c(T,T)
#     ),
#     
#     col.legend = tm_legend(
#       title = "RPI vs. RUE difference",
#       reverse = TRUE,
#       width = 10,
#       height = 30
#     )
#   ) +
#   borders
# 
# rpi_lgs_diff_map <- fill +
#   tm_shape(rpi_lgs_diff) +
#   tm_raster(
#     col.scale = tm_scale_continuous(
#       limits = c(-2, 2),
#       midpoint = 0,
#       values = "RdBu",
#       outliers.trunc = c(T,T)
#     ),
#     
#     col.legend = tm_legend(
#       title = "RPI vs. LGS difference",
#       reverse = TRUE,
#       width = 10,
#       height = 30
#     )
#   ) +
#   borders
# 
# rpi_rue_diff_map
# rpi_lgs_diff_map


```


A more intuitive map might be to plot the quantiles for each dataset, i.e.,
which quantile does each fit into in terms of overall productivity. That is a
scale-free metric.

```{r comparison_maps_quantiles}

rast_as_quantiles <- function(x, quantile_list) {
  
  x_values <- values(x)
  quantiles <- quantile(x_values, quantile_list, na.rm = TRUE)
  
  quantile_values <- findInterval(x_values, quantiles)
  
  out <- quantile_values %>%
    matrix(nrow = nrow(x), ncol = ncol(x), byrow = TRUE) %>%
    rast(extent = ext(x), crs = crs(x))
  
  out
}

quantiles <- seq(0, 1, 0.01)

rpi_quantile_rast <- rast_as_quantiles(rpi_mean, quantiles)
rue_quantile_rast <- rast_as_quantiles(rue_mean, quantiles)
lgs_quantile_rast <- rast_as_quantiles(lgs_100$lgs_ratio, quantiles)

plot(rpi_quantile_rast)
plot(rue_quantile_rast)
plot(lgs_quantile_rast)

# Plot quantile difference maps

# Basemap elements
fill <- tm_shape(ke_tz) +
  tm_fill("grey95")

borders <- tm_shape(ke_tz) +
  tm_borders() +
  tm_layout(frame = FALSE)

rpi_rue_diff_map_q <- fill +
  tm_shape(rpi_quantile_rast - rue_quantile_rast) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(-60, 60),
      midpoint = 0,
      values = "PuOr",
      outliers.trunc = c(TRUE, TRUE)
    ),
    col.legend = tm_legend(
      reverse = TRUE,
      title = "RPI - RUE\n(quantiles)",
      width = 10,
      height = 30
    )
  ) +
  borders

rpi_lgs_diff_map_q <- fill +
  tm_shape(rpi_quantile_rast - lgs_quantile_rast) +
  tm_raster(
    col.scale = tm_scale_continuous(
      limits = c(-60, 60),
      midpoint = 0,
      values = "PuOr",
      outliers.trunc = c(TRUE, TRUE)
    ),
    col.legend = tm_legend(
      reverse = TRUE,
      title = "RPI - LGS\n(quantiles)",
      width = 10,
      height = 30
    )
  ) +
  borders

rpi_rue_diff_map_q
rpi_lgs_diff_map_q


```


We can also plot a sample of points from each method to explore the
quantitative relationship between them.

```{r spatial_comparison_plots}

set.seed(999)
mean_rpi_sample <- spatSample(rpi_mean, 2000, as.points = TRUE,
                              na.rm = TRUE, exhaustive = TRUE)

mean_rue_sample <- extract(rue_mean, mean_rpi_sample)

mean_lgs_100_sample <- extract(lgs_100$lgs_ratio, mean_rpi_sample) %>%
  rename(lgs_100 = lgs_ratio)
mean_lgs_1000_sample <- extract(lgs_1000$lgs_ratio, mean_rpi_sample) %>%
  rename(lgs_1000 = lgs_ratio)

covariate_sample <- extract(covariates, mean_rpi_sample)

spatial_comparison <- mean_rpi_sample %>%
  st_as_sf() %>%
  rename(mean_rpi = mean) %>%
  bind_cols(mean_rue_sample) %>%
  rename(mean_rue = mean) %>%
  bind_cols(mean_lgs_100_sample) %>%
  bind_cols(mean_lgs_1000_sample) %>%
  bind_cols(covariate_sample) %>%
  select(-starts_with("ID"))


ggplot(spatial_comparison, aes(x = mean_rpi, y = mean_rue)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()

ggplot(spatial_comparison, aes(x = mean_rpi, y = lgs_1000)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()

spatial_comparison_valid <- drop_na(spatial_comparison)

cor(spatial_comparison_valid$mean_rpi, spatial_comparison_valid$mean_rue)
cor(spatial_comparison_valid$mean_rpi, spatial_comparison_valid$lgs_100)
cor(spatial_comparison_valid$mean_rpi, spatial_comparison_valid$lgs_1000)

# Results relative to mean precipitation
spatial_comparison %>%
  pivot_longer(1:4, names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = ppt_mean, y = value)) +
  geom_point(size = 0.25) +
  geom_smooth(method = "lm", se = F) +
  facet_wrap(~metric, scales = "free") +
  theme_bw() +
  labs(x = "Mean annual\nprecipitation (mm)",
       y = "Metric")

spatial_comparison %>%
  st_drop_geometry %>%
  pivot_longer(1:4, names_to = "metric", values_to = "value") %>%
  group_by(metric) %>%
  drop_na() %>%
  summarise(cor = cor(ppt_mean, value))


```


We can compare the performance of our method with those of others on various
criteria. For mean values across space, the key issues are:
- Explaining a greater share of spatial variation with environmental
covariates (R-squared)
- Reduced overall variance?
- Consistency (lack of sharp boundaries)
- Lack of correlation with MAP or other key vars

For trend, we can assess performance through:
- High R-squared for each pixel
- Low mean absolute error (weighted?) for each pixel
- Consistency in Mann-Kendall trend?
- P-values of Theil-Sen slope or permutation trend?
