---
title: "Mapping relative productivity index (RPI)"
output: html_notebook
author: Guy Lomax
date: 2023-08-18
---

This Notebook uses quantile regression models fitted on sample data to map
the annual relative productivity index across Kenyan and Tanzanian rangelands
from 2000 to 2018.

TO DO:
1. Fix input layers to filter only to rangeland pixels
2. Include % woody layer to allow fitting of correct model


```{r setup, include = FALSE}

# Data handling
library(tidyverse)
library(sf)
library(terra)
library(here)

# Analysis
library(qgam)
library(gratia)
library(parallel)
library(furrr)
library(tictoc)
library(beepr)

# Visualisation
library(tmap)

tmap_options(check.and.fix = TRUE)

# Set up parallel computing
nc <- detectCores() - 1
plan(multisession, workers = nc)


```


```{r load, include = FALSE}

# Country boundaries
countries <- st_read(here("data", "raw", "vector", "natural_earth",
                          "ne_110m_admin_0_countries_fixed.shp"))

ke_tz <- countries %>% filter(NAME %in% c("Kenya", "Tanzania"))

# Covariate layers
# Static
static_covariates <- rast(here("data", "raw", "raster", "covariateMaps",
                                 "staticVars.tif"))

twi <- rast(here("data", "processed", "raster", "hydroSHEDS",
                 "hydrosheds_twi_fd8.tif"))

# Load dynamic covariates as a named list of rasters (one per year)
dynamic_covariates_paths <- Sys.glob(here("data", "raw", "raster", "covariateMaps",
                                          "dynamicVars*.tif"))

years <- str_extract(dynamic_covariates_paths, "\\d\\d\\d\\d")
dynamic_covariates <- map(dynamic_covariates_paths, rast)
names(dynamic_covariates) <- years

# Fitted models

models <- read_rds(here("data", "processed", "rds", "binned_gam.rds"))

```

Data preparation

```{r data_prep}

# Reproject TWI to same resolution and CRS as other layers

twi_reproj <- project(twi, static_covariates)

# Crop twi and static covariates to dynamic covariates layers
# (GEE doesn't match extents properly due to resampling)

twi_crop <- crop(twi_reproj, dynamic_covariates[[1]])
static_crop <- crop(static_covariates, dynamic_covariates[[1]])

```
Prepare functions for data preparation, model fitting and prediction to raster

```{r functions}

# Convert raster layers to data frame

raster_to_df <- function(dynamic_rast, year) {
  
  # Progress message
  message(paste0("Processing year ", year))
  
  # Extract and join rasters
  all_covariates <- c(dynamic_rast, static_crop, twi_crop)
  
  # Mask to Kenya/Tanzania
  all_covariates_masked <- mask(all_covariates, ke_tz)
  
  # Convert raster to data frame
  covariate_df <- as.data.frame(all_covariates_masked, xy = TRUE, cells = TRUE)
  
  # Add year information
  covariate_df$year <- as.numeric(year)
  
  covariate_df
}

# Prepare data frame variables for prediction

clean_df <- function(df) {
  df_clean <- df %>%
    filter(!is.na(GPP)) %>%
    rename(twi = hydrosheds_twi_fd8,
           parMean = GMT_0900_PAR) %>%
    mutate(woody = shrubland + trees,
           woody_bin = cut_interval(woody, 5))
  
  df_clean
}

# Group rows by woody cover and attach prediction model

attach_model <- function(df, model_df) {
  
  df_joined <- df %>%
    group_by(woody_bin) %>%
    nest() %>%
    left_join(model_df, by = c("woody_bin" = "bin"))
  
  df_joined
}

# Predict potential GPP values
predict_gpp <- function(df) {
  
  df_fitted <- df %>%
    mutate(fitted = future_map2(model, data, predict.gam)) %>%
    mutate(data = map2(data, fitted, cbind)) %>%
    select(-fitted, -model) %>%
    unnest(data) %>%
    ungroup() %>%
    rename("fitted" = ".y[[i]]")

  df_fitted
}

# Calculate RPI

calc_rpi <- function(df) {
  df_rpi <- mutate(df, rpi = GPP / fitted)
  df_rpi
}

# Extract fitted GPP and RPI values as raster layers

df_to_raster <- function(df, target) {
  df_to_convert <- df %>%
    select(x, y, GPP, fitted, rpi)
  
  names <- colnames(df_to_convert)[3:ncol(df_to_convert)]
  
  rpi_raster <- rast(df_to_convert,
                     crs = crs(target),
                     extent = ext(target))
  
  rpi_raster
}


```


```{r calc_rpi}

for (i in seq_len(length(dynamic_covariates))) {
  tic()
  year <- names(dynamic_covariates[i])
  
  dynamic_rast <- dynamic_covariates[[i]]
  
  # Convert to df
  df <- raster_to_df(dynamic_rast, year)
  
  message("Conversion to data frame complete")
  
  # Prepare for prediction
  df_for_prediction <- df %>%
    clean_df() %>%
    attach_model(models)
  
  message("Data frame preparation complete")
  message("Predicting potential GPP from model...")
  
  # Predict potential GPP values from model
  df_fitted <- predict_gpp(df_for_prediction)
  
  message("Model prediction complete")
  
  # Calculate RPI
  
  df_rpi <- calc_rpi(df_fitted)
  
  # Convert back to raster
  
  rpi_rast <- df_to_raster(df_rpi, dynamic_rast)
  
  message("Conversion to raster complete")
  
  # Write to raster file
  filename <- paste0("rpi_rast_", year, ".tif")
  writeRaster(rpi_rast,
              here("data", "processed", "raster", "rpi", filename),
              overwrite = TRUE)
  
  message("Raster written to file: ", filename)
  toc()
  
  # Clear memory for next loop
  rm(dynamic_rast, df, df_for_prediction, df_fitted, df_rpi, rpi_rast)
  gc()
}


```
Apply to full study area to map annual RPI

```{r rpi_map}

rpi_paths <- Sys.glob(here("data", "processed", "raster", "rpi", "rpi_rast_*.tif"))

rpi_list <- map(rpi_paths, rast)

rpi_rast <- map(rpi_list, function(r) {r$rpi}) %>%
  rast()

names(rpi_rast) <- names(dynamic_covariates)

rpi_anim <- tm_shape(ke_tz) + tm_borders() +
  tm_shape(rpi_rast) +
  tm_raster(palette = "PuOr", style = "cont",
            breaks = seq(0, 2,), midpoint = 1) +
  tm_facets(nrow = 1, ncol = 1)

# Save - takes about 5m
tmap_animation(rpi_anim,
  filename = here("results", "figures", "rpi_anim.gif"),
  width = 800, height = 800, dpi = 200)


# Save six years as facets

rpi_facet <- tm_shape(ke_tz) + tm_borders() +
  tm_shape(rpi_rast[[8:19]]) +
  tm_raster(palette = "BrBG", style = "cont",
            breaks = seq(0, 2,), midpoint = 1,
            legend.show = FALSE) +
  tm_facets(nrow = 2, ncol = 6)

tmap_save(rpi_facet,
          here("results", "figures", "rpi_map_facets.png"),
          width = 24, height = 12, units = "cm", dpi = 200)

```

