---
title: "Mapping relative productivity index (RPI)"
output: html_notebook
author: Guy Lomax
date: 2023-08-18
---

This Notebook uses quantile regression models fitted on sample data to map
the annual relative productivity index across Kenyan and Tanzanian rangelands
from 2000 to 2018.


```{r setup, include = FALSE}

# Data handling
library(tidyverse)
library(sf)
library(terra)
library(here)

# Analysis
library(mlr3)
library(ranger)
library(furrr)
library(tictoc)
library(beepr)

# Visualisation
library(tmap)

tmap_options(check.and.fix = TRUE)



```


```{r load, include = FALSE}

# Country boundaries
ke_tz <- st_read(here("data", "raw", "vector", "kenya_tanzania.geojson"))

# Covariate layers
# Static
static_covariates <- rast(here("data", "raw", "raster", "covariateMaps",
                                 "staticVars.tif"))

twi <- rast(here("data", "processed", "raster", "hydroSHEDS",
                 "hydrosheds_twi_fd8.tif"))

names(twi) <- "twi"

# Load dynamic covariates as a named list of rasters (one per year)
dynamic_covariates_paths <- Sys.glob(here("data", "raw", "raster", "covariateMaps",
                                          "dynamicVars*.tif"))

years <- str_extract(dynamic_covariates_paths, "\\d\\d\\d\\d")
dynamic_covariates <- map(dynamic_covariates_paths, rast)
names(dynamic_covariates) <- years

# Load precipitation covariates as a named list of rasters (one per year)
ppt_covariate_paths <- Sys.glob(here("data", "raw", "raster", "covariateMaps",
                                      "pptVars*.tif"))
ppt_covariates <- map(ppt_covariate_paths, rast)
names(ppt_covariates) <- years

# Fitted models

spt_model_tuned <- read_rds(here("results", "rds", "rf_tuned_spt.rds"))
model <- spt_model_tuned$learner$model

```

Data preparation

```{r data_prep}

# Reproject TWI to same resolution and CRS as other layers

twi_reproj <- project(twi, static_covariates)

# Crop covariates to common extent
# (GEE doesn't match extents properly due to resampling)

twi_crop <- crop(twi_reproj, dynamic_covariates[[1]])
static_crop <- crop(static_covariates, dynamic_covariates[[1]])
ppt_covariates_crop <- map(ppt_covariates, crop, y = dynamic_covariates[[1]])


# Calculate mean ppt and mean of mean ppt day layers
mean_ppt <- map(dynamic_covariates, subset, subset = "precipitation") %>%
  rast() %>%
  mean()

mean_meanPptDay <- map(ppt_covariates_crop, subset, subset = "meanPptDay") %>%
  rast() %>%
  mean()

names(mean_ppt) <- "mean_ppt"
names(mean_meanPptDay) <- "mean_meanPptDay"

```


Prepare functions for data preparation, model fitting and prediction to raster

```{r functions}

# Combine raster layers and convert to data frame

raster_to_df <- function(year) {
  
  # Extract yearly covariates
  dynamic_rast <- dynamic_covariates[[year]]
  ppt_rast <- ppt_covariates_crop[[year]]
  
  # Extract and join rasters
  all_covariates <- c(dynamic_rast, ppt_rast,
                      mean_ppt, mean_meanPptDay,
                      static_crop, twi_crop)
  
  # Mask to Kenya/Tanzania
  all_covariates_masked <- mask(all_covariates, ke_tz)
  
  # Convert raster to data frame
  covariate_df <- as.data.frame(all_covariates_masked, xy = TRUE, cells = TRUE)
  
  # Add year information
  covariate_df$year <- as.numeric(year)
  
  covariate_df
}

# Prepare data frame variables for prediction

clean_df <- function(df) {
  df_clean <- df %>%
    filter(!is.na(GPP)) %>%
    remove_missing()
  
  df_clean
}

# Add derived variables: ppt_mean, ppt_anomaly and meanPptDay_anomaly

add_derived_variables <- function(df) {
  df_derived <- df %>%
    mutate(ppt_anomaly = precipitation / mean_ppt,
           meanPptDay_anomaly = meanPptDay - mean(meanPptDay)) %>%
    rename(sdii = intensity)
  
  df_derived
}

# Predict potential GPP values
predict_gpp <- function(df) {
  
  predictions_q <- predict(model, data = df, type = "quantiles", quantiles = 0.9)
  predictions_m <- predict(model, data = df)
  
  df_fitted <- df %>%
    mutate(quantile_pred = predictions_q$predictions,
           mean_pred = predictions_m$predictions)
  
  df_fitted
}

# Calculate RPI

calc_rpi <- function(df) {
  df_rpi <- mutate(df, rpi = GPP / quantile_pred)
  df_rpi
}

# Extract fitted GPP and RPI values as raster layers

df_to_raster <- function(df, target) {
  df_to_convert <- df %>%
    select(x, y, GPP, quantile_pred, mean_pred, rpi)
  
  names <- colnames(df_to_convert)[3:ncol(df_to_convert)]
  
  rpi_raster <- rast(df_to_convert,
                     crs = crs(target),
                     extent = ext(target))
  
  rpi_raster
}


```


Loop through annual raster layers to create maps of predicted potential GPP and
RPI. Using a loop to reduce memory usage and ensure years are written to disk
as processing is completed.

```{r calc_rpi, eval = FALSE}

for (i in years) {
  tic()
  
  # Progress message
  message(paste0("Processing year ", i))

  # Convert to df
  df <- raster_to_df(i)
  
  message("Conversion to data frame complete")
  
  # Prepare for prediction
  df_for_prediction <- df %>%
    clean_df() %>%
    add_derived_variables()
  
  message("Data frame preparation complete")
  message("Predicting potential GPP from model...")
  
  # Predict potential GPP values from model
  df_fitted <- predict_gpp(df_for_prediction)
  
  message("Model prediction complete")
  
  # Calculate RPI
  
  df_rpi <- calc_rpi(df_fitted)
  
  # Convert back to raster
  target_rast <- dynamic_covariates[[i]]
  rpi_rast <- df_to_raster(df_rpi, target_rast)
  
  message("Conversion to raster complete")
  
  # Write to raster file
  filename <- paste0("rf_rpi_", i, ".tif")
  writeRaster(rpi_rast,
              here("data", "processed", "raster", "rpi", filename),
              overwrite = TRUE)
  
  message("Raster written to file: ", filename)
  toc()
  
  # Clear memory for next loop
  rm(df, df_for_prediction, df_fitted, df_rpi, target_rast, rpi_rast)
  gc()
}



```


Apply to full study area to map annual RPI

```{r rpi_map}

# Load exported RPI annual rasters
rpi_paths <- Sys.glob(here("data", "processed", "raster", "rpi", "rf_rpi_rast*.tif"))

rpi_list <- map(rpi_paths, rast)

rpi_rast <- map(rpi_list, function(r) {r$rpi}) %>%
  rast()

names(rpi_rast) <- years

# Calculate mean and variance

rpi_rast_mean <- mean(rpi_rast)
rpi_rast_var <- app(rpi_rast, var, na.rm = T)

writeRaster(rpi_rast_mean,
            here("data", "processed", "raster", "rpi", "rpi_rast_mean.tif"))

# mean_map <- tm_shape(ke_tz) + tm_fill(col = "grey85") +
#   tm_shape(rpi_rast_mean) +
#   tm_raster(col.scale = tm_scale_continuous(
#     limits = c(0,1.6), midpoint = 0.63, values = "RdBu"
#   )) +
#   tm_shape(ke_tz) +
#   tm_borders()

# # Save six years as facets
# 
# rpi_facet <- tm_shape(ke_tz) + tm_fill(col = "grey85") +
#   tm_shape(rpi_rast[[1:12]]) +
#   tm_raster(
#     col.scale = tm_scale_continuous(limits = c(0,1.6), midpoint = 0.63, values = "RdBu"),
#     col.legend = tm_legend(show = FALSE)) +
#   tm_shape(ke_tz) +
#   tm_borders() +
#   tm_facets_wrap(nrow = 2, ncol = 6)
# 
# tmap_save(rpi_facet,
#           here("results", "figures", "rpi_map_facets_with_time.png"),
#           width = 24, height = 12, units = "cm", dpi = 200)

```


There are some clear patterns in the mean values, but annual values still
vary widely with no clear stable patterns. I will need to work with mean
values, trends and persistently low/high-value pixels.

Regarding mean values, the Maasai Mara and Serengeti appear to have
above-average values, as do North-East Kenya, some parts of Northern and Central
Kenya and Eastern and Southern Tanzania. It would be interesting to see whether
these correspond to particular parks, protected areas etc.

In addition. very high values can be seen close to rivers, lakes or wetlands
(e.g., the Lorian Swamp is clearly visible as a high-RPI area). Some of the
high-RPI areas also seem to be associated with high tree-cover areas (e.g.,
North-East Kenya and Southern Tanzania), but not all. North-West Kenya (Lake
Turkana region) has some very low values, which might represent issues in the
PML_V2 dataset (since much of that area is classified as "barren land" with a
GPP of zero by definition).

```{r persistence_and_trends}

# Trend (using Theil-Sen)

fit_theil_sen <- function(vector) {
  
  if (any(is.na(vector))) {
    rep(NA, 4)
  }
  
  else {
    regression_df <- tibble(years = seq_len(length(vector)),
                            rpi = vector)
    
    ts_regression <- RobustLinearReg::theil_sen_regression(rpi ~ years, regression_df)
    
    slope_params <- broom::tidy(ts_regression)[2,2:5] %>%
      as.vector() %>%
      unlist()
    
    slope_params
  }
}

# rpi_ts_trend <- app(rpi_rast, fit_theil_sen)
# names(rpi_ts_trend) <- c("slope", "std_error", "f_statistic", "p_value")
# 
# writeRaster(rpi_ts_trend,
#             here("results", "figures", "rf_rpi_ts_trend_rast.tif"),
#             overwrite = TRUE)

rpi_ts_trend <- rast(here("data", "processed", "raster", "rpi", "rf_rpi_ts_trend_rast.tif"))

# Trend (using Mann Kendall test)

calculate_kendall_tau <- function(vector) {
  
  if (any(is.na(vector))) {
    rep(NA, 2)
  }
  
  else {
    
    kendall_test <- unlist(Kendall::MannKendall(vector))
    
    kendall_test[1:2]
  }
}

# rpi_kendall_tau <- app(rpi_rast, calculate_kendall_tau, cores = 32)
# names(rpi_kendall_tau) <- c("tau", "p_value")
# 
# writeRaster(rpi_kendall_tau,
#             here("results", "figures", "rf_rpi_kendall_tau_rast.tif"),
#             overwrite = TRUE)

rpi_kendall_tau <- rast(here("data", "processed", "raster", "rpi", "rf_rpi_kendall_tau_rast.tif"))


## Visualise in tmap viewer
tmap_mode("view")

tm_shape(rpi_rast_mean) +
  tm_raster(col.scale = tm_scale_continuous(
    limits = c(0,1.6),
    midpoint = 0.63,
    values = "RdBu",
    outliers.trunc = c(T,T))) +
  tm_shape(rpi_ts_trend) +
  tm_raster(col = "slope",
            col.scale = tm_scale_continuous(
              limits = c(-0.02, 0.02),
              midpoint = 0,
              values = "RdBu",
              outliers.trunc = c(T,T))) +
  tm_shape(rpi_kendall_tau$tau) +
  tm_raster(col.scale = tm_scale_continuous(
    limits = c(-0.6, 0.6),
    midpoint = 0,
    values = "RdBu",
    outliers.trunc = c(T,T))) +
  tm_basemap(c("OpenStreetMap", "Esri.WorldImagery")) + tm_scalebar()

```


Identify persistently low- or high-valued pixels

``` {r persistent_low_high}

# Find pixels with mean RPI < 0.5 (or < median) and trend 0 +/- 0.005

mean_quantiles <- quantile(values(rpi_rast_mean), c(0.1, 0.2, 0.333, 0.5, 0.666, 0.8, 0.9), na.rm = TRUE)
trend_quantiles <- quantile(values(rpi_ts_trend$slope), c(0.1, 0.2, 0.333, 0.5, 0.666, 0.8, 0.9), na.rm = TRUE)

low_rpi <- rpi_rast_mean <= mean_quantiles[1] & rpi_ts_trend$slope >= trend_quantiles[3] & rpi_ts_trend$slope <= trend_quantiles[5]
high_rpi <- rpi_rast_mean >= mean_quantiles[7] & rpi_ts_trend$slope >= trend_quantiles[3] & rpi_ts_trend$slope <= trend_quantiles[5]
low_trend <- rpi_ts_trend$slope <= trend_quantiles[1]
high_trend <- rpi_ts_trend$slope >= trend_quantiles[7]

rpi_cats <- (low_rpi + 2 * high_rpi + 3 * low_trend + 4 * high_trend)
rpi_cats_masked <- mask(rpi_cats, rpi_cats, maskvalues = c(0,NA))


tm_shape(rpi_rast_mean) +
  tm_raster(col.scale = tm_scale_continuous(
    limits = c(0,1.6),
    midpoint = 0.63,
    values = "RdBu",
    outliers.trunc = c(T,T))) +
  tm_shape(rpi_ts_trend) +
  tm_raster(col = "slope",
            col.scale = tm_scale_continuous(
              limits = c(-0.02, 0.02),
              midpoint = 0,
              values = "RdBu",
              outliers.trunc = c(T,T))) +
  tm_shape(rpi_cats_masked) +
  tm_raster(col.scale = tm_scale_categorical(values = c('brown','cornflowerblue','orange','purple'))) +
  tm_basemap(c("OpenStreetMap", "Esri.WorldImagery")) + tm_scalebar()

# Find pixels with mean RPI > 0.8 and trend 0 +/- 0.005

# Find pixels with high (AR1)? Or negative AR(1)?? Or just high variance and
# low trend



```


Subset by % tree cover and/or add to map as additional layer/mask

```{r rpi_tc}

tree_cover <- static_covariates$wriTCFrac

tm_shape(tree_cover) + tm_raster(col.scale = tm_scale(limits = c(0,1), values = "Greens")) +
  tm_shape(rpi_rast_mean) +
  tm_raster(col.scale = tm_scale_continuous(
    limits = c(0,1.6),
    midpoint = 0.63,
    values = "RdBu",
    outliers.trunc = c(T,T))) +
  tm_basemap(c("OpenStreetMap", "Esri.WorldImagery")) + tm_scalebar()

# Combine into a raster and sample

# tree_cover_rpi <- tree_cover %>%
#   crop(rpi_rast_mean) %>%
#   c(rpi_rast_mean, rpi_ts_trend$slope)
# 
# set.seed(999)
# tree_cover_rpi_sample <- spatSample(tree_cover_rpi, 2000, xy = TRUE, na.rm = TRUE)
# 
# head(tree_cover_rpi_sample)
# 
# cor(tree_cover_rpi_sample$wriTCFrac, tree_cover_rpi_sample$mean)
# 
# ggplot(tree_cover_rpi_sample, aes(x = wriTCFrac, y = slope)) +
#   geom_point(size = 0.2, alpha = 0.2) +
#   geom_smooth()

# Create a polygon map of different bins
# Bins should be 30-60% and >60%, or 25-50, 50-75 and 75-100%

tree_cover_bin1 <- tree_cover > 0.3 & tree_cover <= 0.6
tree_cover_bin2 <- (tree_cover > 0.6) * 2

tree_cover_bins <- tree_cover_bin1 + tree_cover_bin2

tm_shape(tree_cover) +tm_raster(col.scale = tm_scale(limits = c(0,1), values = "Greens")) +
  tm_shape(tree_cover_bins) + tm_raster()


# Too varied to meaningfully polygonise - I'll simply need to create masks
opaque_tc_mask <- tree_cover_bins
values(opaque_tc_mask)[values(opaque_tc_mask) == 0] <- NA
opaque_tc_mask <- crop(opaque_tc_mask, rpi_rast_mean)

tm_shape(rpi_rast_mean) +
  tm_raster(col.scale = tm_scale_continuous(
    limits = c(0,1.6),
    midpoint = 0.63,
    values = "RdBu",
    outliers.trunc = c(T,T))) +
  tm_shape(rpi_ts_trend) +
  tm_raster(col = "slope",
            col.scale = tm_scale_continuous(
              limits = c(-0.02, 0.02),
              midpoint = 0,
              values = "RdBu",
              outliers.trunc = c(T,T))) +
  tm_shape(rpi_cats_masked) +
  tm_raster(col.scale = tm_scale_categorical(values = c('brown','cornflowerblue','orange','purple'))) +
  tm_shape(opaque_tc_mask) +
  tm_raster(palette = "black", alpha = 0.75) +
  tm_basemap(c("OpenStreetMap", "Esri.WorldImagery")) + tm_scalebar()

```


Now let's use the protected area dataset to test mean RPI in particular parks,
reserves or community conservancies, with and without stratification by
tree cover.

```{r rpi_protected_areas}

protected_areas <- st_read(here("data", "raw", "vector", "protected_areas",
                                "terrestrialPAsKeTz.geojson"))

unique(protected_areas$DESIG_ENG)

target_pas <- c("National Park", "Community Conservancy", "National Reserve",
                "Community Nature Reserve", "Community Conservancy", "Game Reserve")

protected_areas_target <- protected_areas %>%
  filter(DESIG_ENG %in% target_pas) %>%
  # filter(PARENT_ISO == "KEN") %>%
  filter(STATUS != "Proposed")

protected_areas_rpi <- protected_areas_target %>%
  mutate(mean_rpi = zonal(rpi_rast_mean, vect(protected_areas_target), na.rm = TRUE)$mean)

protected_areas_rpi %>%
  filter(DESIG_ENG == "National Park" & ISO3 == "KEN") %>%
  ggplot(aes(x = NAME, y = mean_rpi)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45))

```

